# Reproducible Research: Peer Assessment 1

The instructions for this  assignment can be found on this page:
[https://github.com/rdpeng/RepData_PeerAssessment1](https://github.com/rdpeng/RepData_PeerAssessment1).  
Please read this first for context and details.

Calculated with R 3.1.0, for details see bottom of report.


## Loading and preprocessing the data

- Set working directory, 
- Read in data, ignoring cases with "NA" values
- Converting the "date" column to Date datatype

```{r}

setwd("/home/knut/Documents/coursera/datascience/5-reprod-research/RepData_PeerAssessment1")

act <- read.csv("activity.csv", header=TRUE, na.strings=c("NA"))


actf <- act[which(act$steps >= 0),]
actf$date <- as.Date(actf$date)


```

 These are the number of NA values in each column:
```{r}
colSums(is.na(act));nas_per_column
```



### The mean of steps per each day can be calculated as follows:
```{r}
library(plyr)

actfagg <- ddply(actf, .(date), summarize,  sumsteps = round(sum(steps), 0))
head(actfagg, 10)

```
### The mean and median total number of steps taken per day are very similar.
Mean:
```{r}

mean_steps_per_day <- round(mean(actfagg$sumsteps),0); mean_steps_per_day

```

Median:
```{r}
median_steps_per_day <- round(median(actfagg$sumsteps),0); median_steps_per_day
```
### Steps per day visualised as a histogram:
```{r hist}

hist(actfagg$sumsteps,  breaks=20, xlab="Number of steps walked per day", ylab="Frequency of days", main="Daily physical activity of an anonymous person in 2012")


```




## Are there differences in activity patterns between weekdays and weekends?

To answer  this question, we add a new column to the filtered `act` data frame.  
This column will contain the weekday encoded as a number.

```{r}

library(lubridate)
act$wkday <- wday(act$date)

```


# R Session Info:

```{r}
sessionInfo()
```